---
title: "Structural Code Violation Analysis in Philadelphia"
author: "Manas Vesvikar"
date: "2025-06-01"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(tidymodels)
library(janitor)
library(lubridate)
library(ggplot2)
library(vip)
library(rpart.plot)
library(naniar)
```

#  Introduction

This report investigates structural code violations in the city of Philadelphia using a 5K-sample dataset of property code violations. The objective is to build interpretable models to classify whether a violation is structural in nature, based on categorical and spatial attributes.

#  Data Overview

The dataset includes 5,000 code violations with variables such as case status, violation status, council district, latitude/longitude, and a title describing the violation. We cleaned and transformed the data to create a binary outcome variable, `is_structural_violation`.

```{r load-data}
violations <- read_csv("data_processed/violations_sample.csv") %>%
  clean_names()

head(violations)
```

##  Missing Data Overview

```{r missing-data-visuals}
gg_miss_var(violations) + labs(title = "Missing Data by Column (Sample Dataset)")

vis_miss(violations[1:500, ]) +
  labs(title = "Missing Data Heatmap (First 500 Rows)")
```

#  Exploratory Data Analysis

##  Monthly Violation Trends

```{r monthly-trends}
violations %>%
  mutate(casecreateddate = ymd_hms(casecreateddate),  # Ensure proper datetime parsing
         month = floor_date(casecreateddate, "month")) %>%
  filter(!is.na(month)) %>%  # Remove rows where date parsing failed
  count(month) %>%
  ggplot(aes(x = month, y = n)) +
  geom_line(color = "brown", size = 1) +
  labs(title = "Monthly Violation Trends",
       x = "Month", y = "Number of Violations") +
  theme_minimal()
```

##  Violation Status Distribution

```{r status-distribution}
violations %>%
  count(violationstatus) %>%
  ggplot(aes(x = fct_reorder(violationstatus, n), y = n, fill = violationstatus)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Violation Status Distribution", x = "Status", y = "Count") +
  theme_minimal()
```

##  Top 10 Violation Types

```{r top-violations}
violations %>%
  count(violationcodetitle, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = fct_reorder(violationcodetitle, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Violation Types", x = "Violation Title", y = "Count") +
  theme_minimal()
```

#  Modeling

We compare two models: logistic regression and a decision tree.

```{r load-train-test}
train_data <- read_csv("data_processed/train_data.csv")
test_data <- read_csv("data_processed/test_data.csv")
```

##  Logistic Regression

```{r logreg-model}
log_recipe <- recipe(is_structural_violation ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>% 
  step_normalize(all_numeric_predictors())

log_spec <- logistic_reg() %>%
  set_engine("glm")

log_wf <- workflow() %>%
  add_recipe(log_recipe) %>%
  add_model(log_spec)

log_fit <- fit(log_wf, data = train_data)

log_preds <- predict(log_fit, test_data, type = "prob") %>%
  bind_cols(predict(log_fit, test_data)) %>%
  bind_cols(test_data %>% select(is_structural_violation)) %>%
  mutate(is_structural_violation = factor(is_structural_violation))

log_metrics <- metric_set(accuracy, roc_auc, kap)(
  log_preds,
  truth = is_structural_violation,
  estimate = .pred_class,
  .pred_Yes
)
log_metrics
```

##  Decision Tree

```{r tree-model}
tree_recipe <- recipe(is_structural_violation ~ ., data = train_data)
tree_spec <- decision_tree(tree_depth = 5) %>%
  set_engine("rpart") %>%
  set_mode("classification")

tree_wf <- workflow() %>%
  add_recipe(tree_recipe) %>%
  add_model(tree_spec)

tree_fit <- fit(tree_wf, data = train_data)

tree_preds <- predict(tree_fit, test_data, type = "prob") %>%
  bind_cols(predict(tree_fit, test_data)) %>%
  bind_cols(test_data %>% select(is_structural_violation)) %>%
  mutate(is_structural_violation = factor(is_structural_violation))

tree_metrics <- metric_set(accuracy, roc_auc, kap)(
  tree_preds,
  truth = is_structural_violation,
  estimate = .pred_class,
  .pred_Yes
)
tree_metrics
```

##  Model Comparison – ROC Curve

```{r roc-curve}
log_roc <- predict(log_fit, test_data, type = "prob") %>%
  bind_cols(test_data %>% mutate(is_structural_violation = factor(is_structural_violation))) %>%
  roc_curve(truth = is_structural_violation, .pred_Yes) %>%
  mutate(model = "Logistic Regression")

tree_roc <- predict(tree_fit, test_data, type = "prob") %>%
  bind_cols(test_data %>% mutate(is_structural_violation = factor(is_structural_violation))) %>%
  roc_curve(truth = is_structural_violation, .pred_Yes) %>%
  mutate(model = "Decision Tree")

bind_rows(log_roc, tree_roc) %>%
  autoplot() +
  labs(title = "ROC Curves: Logistic Regression vs Decision Tree") +
  theme_minimal()
```

##  Decision Tree Visualization

```{r tree-plot}
final_tree <- extract_fit_parsnip(tree_fit)$fit
rpart.plot(final_tree, main = "Decision Tree Structure")
```

##  Feature Importance (Decision Tree)

```{r vip-tree}
final_tree_model <- extract_fit_parsnip(tree_fit)$fit
vip(final_tree_model,
    num_features = 10,
    bar = TRUE,
    aesthetics = list(fill = "forestgreen")) +
  ggtitle("Top Features – Decision Tree Model") +
  theme_minimal()
```

#  Conclusion

Both models achieved the same accuracy (~77.4%), but the decision tree had a slightly higher ROC AUC (0.289 vs. 0.223). The top predictors were `caseprioritydesc` and `violationstatus`. The decision tree provided stronger interpretability through visual structure and VIP analysis, while logistic regression gave smoother probability estimates.

This analysis can support city inspectors and planners in prioritizing inspections based on structural risk signals visible in public case records.

